Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2023-11-14 manuscript/manuscript.Rmd
Further changes.

## Robust shift-sensitive power transformations

@@

Modifications intended to make power transformations invariant to shifts and methods to improve their robustness against outliers need to be assessed using data drawn from a range of different distributions. Since the power transformations are intended for use with unimodal distributions, the generalised normal distribution [@Subbotin1923-qk; @Nadarajah2005-xe] is a suitable option for simulating realistic feature distributions. This distribution has the following probability density function $f_{\beta}$ for a value $x \in \mathbb{R}$:

@@

Let central normality be defined as normality of central portion $\kappa$ of the data, i.e. $\mathbf{X}_{\text{central}} = \left\{x_i \in \mathbf{X} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}$, with $p_i$ probabilities of the empirical distribution, as previously.
We then compute the residual errors between the z-scores of the transformed feature $\phi^{\lambda, x_0} (\mathbf{X})$ and the expected z-scores from a standard normal distribution: $r_i =\left| \left( \phi^{\lambda, x_0}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|$, with $\mu_M$ and $\sigma_M$ are robust Huber M-estimates of location and scale of the transformed feature $\phi^{\lambda, x_0} (\mathbf{X})$ [@Huber1981-su].
The set of residual errors for the central portion of the data is then $\mathbf{R}_{\text{central}} = \left\{ r_i \in \left\{ r_1, r_2, \ldots, r_n\right\} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}$.

@@

## Shift-sensitive transformations are location-invariant

To assess whether the modified power transformations lead to $\lambda$ values that are invariant to shifts, we simulate three different distributions. The first distribution is a normal distribution, as previously shown in Figure \ref{fig:decreased-normality}.
We first randomly draw $10000$ values from a normal distribution: $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)$, or equivalently $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)$.
The second distribution is a right-skewed normal distribution $\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)$.
The third distribution is a left-skewed normal distribution $\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)$.
We then compute transformation parameter $\hat{\lambda}$ using the original definitions (equations \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}) and the shift-sensitive definitions (equations \ref{eqn:box-cox-shift} and \ref{eqn:yeo-johnson-shift}) for each distribution, after adding a positive value $d$ to each distribution. Here, $d \in [1, 10^6]$.

```{r shifted-distributions, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Shift-sensitive power transformation produces transformation parameters that are invariant to location. ",
  "Samples were drawn from normal, right-skewed and left-skewed distributions, respectively,  which then underwent a shift $d$. ",
  "Estimates of transformation parameter $\\lambda$ with the original power transformations (circle) show strong dependency on the overall location of the distribution, ",
  "whereas estimates with the shift-sensitive power transformations (triangle) are constant.")

@@

The result is shown in Figure \ref{fig:shifted-distributions}. For each distribution, transformation parameter $\hat{\lambda}$ varies with $d$ when estimated using the original definitions. In contrast, estimation of $\lambda$ using shift-sensitive power transformations was invariant to $d$.

@@

# Appendix A: Non-MLE-based optimisation of transformation parameters

@@

Let $\mathbf{X}$ be a feature with ordered feature values, and $\mathbf{Y}^\lambda =\psi^{\lambda} \left(\mathbf{X} \right)$ and $\mathbf{Y}^{\lambda, x_0} =\psi^{\lambda, x_0} \left(\mathbf{X} \right)$ its transformed values using conventional and shift-sensitive power transformations, respectively. Since power transformations are monotonic, $\mathbf{Y}$ will likewise be ordered.

Below we will focus on criteria based on empirical density function, and on skewness and kurtosis of the transformed featured. Other potential criteria, such as the Shapiro-Wilk test statistic [@Shapiro1965-zd] are not investigated here. In the case of the Shapiro-Wilk test statistic this is because of scalability to features with many ($> 5000$) instances, and because adapting the test statistic to include weights is not straightforward.

>>>>

## Robust location and scale invariant power transformations

@@

Modifications intended to make power transformations invariant to location and scale of a feature and methods to improve their robustness against outliers need to be assessed using data drawn from a range of different distributions. Since the power transformations are intended for use with unimodal distributions, the generalised normal distribution [@Subbotin1923-qk; @Nadarajah2005-xe] is a suitable option for simulating realistic feature distributions. This distribution has the following probability density function $f_{\beta}$ for a value $x \in \mathbb{R}$:

@@

Let central normality be defined as normality of central portion $\kappa$ of the data, i.e. $\mathbf{X}_{\text{central}} = \left\{x_i \in \mathbf{X} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}$, with $p_i$ probabilities of the empirical distribution, as previously.
We then compute the residual errors between the z-scores of the transformed feature $\phi^{\lambda, x_0, s} (\mathbf{X})$ and the expected z-scores from a standard normal distribution: $r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|$, with $\mu_M$ and $\sigma_M$ robust Huber M-estimates of location and scale of the transformed feature $\phi^{\lambda, x_0, s} (\mathbf{X})$ [@Huber1981-su].
The set of residual errors for the central portion of the data is then $\mathbf{R}_{\text{central}} = \left\{ r_i \in \left\{ r_1, r_2, \ldots, r_n\right\} \, | \,  \frac{1-\kappa}{2} \leq  p_i \leq \frac{1 + \kappa}{2}\right\}$.

@@

## Invariance to location and scale

To assess whether the modified power transformations lead to $\lambda$ values that are invariant to location and scale of the distribution, we simulate three different distributions. The first distribution is a normal distribution, as previously shown in Figure \ref{fig:decreased-normality}.
We first randomly draw $10000$ values from a normal distribution: $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)$, or equivalently $\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)$.
The second distribution is a right-skewed normal distribution $\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)$.
The third distribution is a left-skewed normal distribution $\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)$.
We then compute transformation parameter $\hat{\lambda}$ using the original definitions (equations \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}) and the location and scale invariant definitions (equations \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) for each distribution. To assess location invariance, a positive value $d_{\text{shift}}$ is added to each distribution with $d_{\text{shift}} \in [1, 10^6]$. Similarly, to assess scale invariance, each distribution is multiplied by a positive value $d_{\text{scale}}$, where $d_{\text{scale}} \in [1, 10^6]$.

```{r shifted-distributions, warning=FALSE, message=FALSE, fig.cap=cap}
cap <- paste0(
  "Invariant power transformation produces transformation parameters that are invariant to location and shape. ",
  "Samples were drawn from normal, right-skewed and left-skewed distributions, respectively,  which then underwent a shift $d_{\\text{shift}}$ or multiplication by $d_{\\text{scale}}$. ",
  "Estimates of transformation parameter $\\lambda$ with the conventional power transformations (circle) show strong dependency on the overall location and scale of the distribution, ",
  "whereas estimates with the invariant power transformations (triangle) are constant.")

@@

The result is shown in Figure \ref{fig:shifted-distributions}. For each distribution, transformation parameter $\hat{\lambda}$ varies with $d_{\text{shift}}$ and $d_{\text{scale}}$ when estimated using the original definitions. In contrast, estimation of $\lambda$ using invariant power transformations was invariant to $d_{\text{shift}}$ and $d_{\text{scale}}$.

@@

# Appendix A: Log-likelihood functions for invariant transformations.

Location and scale-invariant Box-Cox and Yeo-Johnson transformations are parameterised using location $x_0$ and scale $s$ parameters, in addition to transformation parameter $\lambda$.
This leads to the following transformations. The location and scale-invariant Box-Cox transformation is:

```{=latex}
\begin{equation}
\phi_{\text{BC}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left(\frac{x_i - x_0}{s} \right)^\lambda - 1 \right) / \lambda & \text{if } \lambda \neq 0\\
\log\left[\frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0
\end{cases}
\end{equation}
```
Where $x_i - x_0 > 0$.
The location and scale-invariant Yeo-Johnson transformation is:

```{=latex}
\begin{equation}
\phi_{\text{YJ}}^{\lambda, x_0, s} (x_i) = 
\begin{cases}
\left( \left( 1 + \frac{x_i - x_0}{s}\right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x_i - x_0 \geq 0\\
\log\left[1 + \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 0 \text{ and } x_i - x_0 \geq 0\\
-\left( \left( 1 - \frac{x_i - x_0}{s}\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x_i - x_0 < 0\\
-\log\left[1 - \frac{x_i - x_0}{s}\right] & \text{if } \lambda = 2 \text{ and } x_i - x_0 < 0
\end{cases}
\end{equation}
```

The parameters of these power transformations can be optimised based by maximising the log-likelihood function, under the assumption that the transformed feature $\phi^{\lambda, x_0, s} (\mathbf{X})$ follows a normal distribution.
The log-likelihood functions for conventional Box-Cox and Yeo-Johnson transformations are well-known.
However, the introduction of scaling parameter $s$ prevents their direct use.
Here, we first derive the general form of the log-likelihood functions, and then derive their power-transformation specific definitions.

Let $f(x_1, \ldots, x_n)$ be the probability density function of feature $\mathbf{X} = \{ x_1, \ldots, x_n\}$, 
and $f^{\lambda, x_0, s} (\phi^{\lambda, x_0, s}(x_1), \ldots, \phi^{\lambda, x_0, s}(x_n))$ be the probability density function of 
the transformed feature $\phi^{\lambda, x_0, s} (\mathbf{X})$, that is assumed to follow a normal distribution.

The two probability density functions are related as follows:

```{=latex}
\begin{equation}
f^{\lambda, x_0, s}(x_1, \ldots, x_n) = f^{\lambda, x_0, s} (\phi^{\lambda, x_0, s}(x_1), \ldots, \phi^{\lambda, x_0, s}(x_n)) \left|\mathbf{J}\right|
\end{equation}
```
Where, $\left|\mathbf{J}\right|$ is the determinant of Jacobian $\mathbf{J}$.
The Jacobian takes the following form, with off-diagonal elements $0$:
```{=latex}
\begin{equation}
\mathbf{J} =
\begin{bmatrix}
    \frac{\partial}{\partial x_1} \phi^{\lambda, x_0, s}(x_1) & 0 & \dots & 0 \\
    0 & \frac{\partial}{\partial x_2} \phi^{\lambda, x_0, s}(x_2) & \dots & 0 \\
    \vdots & \vdots  & \ddots &  \vdots \\
    0  & 0 & 0 & \frac{\partial}{\partial x_n} \phi^{\lambda, x_0, s}(x_n)
\end{bmatrix}
\end{equation}
```

Thus, $\left| \mathbf{J} \right| = \prod_{i=1}^n \frac{\partial}{\partial x_i} \phi^{\lambda, x_0, s}(x_i)$.

Since in our situation $\{x_1, \ldots, x_n\}$ in $f^{\lambda, x_0, s}(x_1, \ldots, x_n)$ are considered fixed (i.e., known), $f^{\lambda, x_0, s}(x_1, \ldots, x_n)$ may be considered a likelihood function.
The log-likelihood function $\mathcal{l}^{\lambda, x_0, s}$ is then:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}^{\lambda, x_0, s} & = \log f^{\lambda, x_0, s}(x_1, \ldots, x_n) \\
 & = \log \left[ f^{\lambda, x_0, s} (\phi^{\lambda, x_0, s}(x_1), \ldots, \phi^{\lambda, x_0, s}(x_n)) \right] + \log \left|\mathbf{J}\right| \\
 & = \log \left[ f^{\lambda, x_0, s} (\phi^{\lambda, x_0, s}(x_1), \ldots, \phi^{\lambda, x_0, s}(x_n)) \right] + \log \prod_{i=1}^n \frac{\partial}{\partial x_i} \phi^{\lambda, x_0, s}(x_i) \\
 & = -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi^{\lambda, x_0, s}(x_i) - \mu \right)^2 + \sum_{i=1}^n \log \left[ \frac{\partial}{\partial x_i} \phi^{\lambda, x_0, s}(x_i)\right]
\end{split}
\end{equation}
```
With $\mu$ the average of $\phi^{\lambda, x_0, s}(\mathbf{X})$ and $\sigma^2$ its variance.
The first two terms derive directly from the log-likelihood function of a normal distribution, and are not specific to the type of power transformation used.
However, the final term differs between Box-Cox and Yeo-Johnson transformations.

## Box-Cox transformation

For the Box-Cox transformation the partial derivative is:
```{=latex}
\begin{equation}
\begin{split}
\frac{\partial}{\partial x_i} \phi_{\text{BC}}^{\lambda, x_0, s}(x_i) & = \frac{1}{s} \left(\frac{x_i - x_0}{s} \right)^{\lambda-1} \\
 & = \frac{1} {s^\lambda} \left(x_i - x_0 \right)^{\lambda - 1}
\end{split}
\end{equation}
```

Thus the final term in $\mathcal{l}_{\text{BC}}^{\lambda, x_0, s}$ is:
```{=latex}
\begin{equation}
\begin{split}
\sum_{i=1}^n \log \frac{\partial}{\partial x_i} \phi_{\text{BC}}^{\lambda, x_0, s}(x_i) & = \sum_{i=1}^n \log \left[ s^{-\lambda} (x_i - x_0)^{\lambda - 1} \right] \\
& = \sum_{i=1}^n \log \left[s^{-\lambda} \right] + \log \left[ (x_i - x_0)^{\lambda - 1} \right]\\
& = -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```

This leads to the following log-likelihood for the Box-Cox transformation:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{BC}}^{\lambda, x_0, s} = & -\frac{n}{2} \log \left[2 \pi \sigma^2 \right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& -n \lambda \log s + \left( \lambda - 1 \right) \sum_{i=1}^n \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```

We introduced per-sample weights $w_i$ to facilitate robust power transformations. The log likelihood of the robust Box-Cox transformation is:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{rBC}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \lambda \left( \sum_{i=1}^n w_i \right) \log s + \left( \lambda - 1 \right) \sum_{i=1}^n w_i \log \left[ x_i - x_0 \right]
\end{split}
\end{equation}
```
where $\mu_w$ and $\sigma^2_w$ are the weighted mean and weighted variance of the Box-Cox transformed feature $\phi_{\text{BC}}^{\lambda, x_0, s} (\mathbf{X})$, respectively:

## Yeo-Johnson transformation

For the Yeo-Johnson transformation, the partial derivative is:
```{=latex}
\begin{equation}
\frac{\partial}{\partial x_i} \phi_{\text{YJ}}^{\lambda, x_0, s}(x_i) =
\begin{cases}
\frac{1}{s} \left(1 + \frac{x_i - x_0}{s}\right)^{\lambda - 1} & \text{if } x_i - x_0 \geq 0\\
\frac{1}{s} \left(1 - \frac{x_i - x_0}{s}\right)^{1 - \lambda} & \text{if } x_i - x_0 < 0
\end{cases}
\end{equation}
```

Thus the final term in $\mathcal{l}_{\text{YJ}}^{\lambda, x_0, s}$ is:
```{=latex}
\begin{equation}
\begin{split}
\sum_{i=1}^n \log \frac{\partial}{\partial x_i} \phi_{\text{YJ}}^{\lambda, x_0, s}(x_i) & = - n \log s + (\lambda - 1) \sum_{i=1}^n \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
```

This lead to the following log-likelihood for the Yeo-Johnson transformation:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{YJ}}^{\lambda, x_0, s} = & -\frac{n}{2} \log\left[2 \pi \sigma^2\right] -\frac{1}{2 \sigma^2} \sum_{i=1}^n \left( \phi^{\lambda, x_0, s}(x_i) - \mu \right)^2 \\
& - n \log s + (\lambda - 1) \sum_{i=1}^n \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
```

The log likelihood of the robust Yeo-Johnson transformation is:
```{=latex}
\begin{equation}
\begin{split}
\mathcal{l}_{\text{rYJ}}^{\lambda, x_0, s} = & -\frac{1}{2} \left(\sum_{i=1}^n w_i \right) \log \left[ 2 \pi \sigma_w^2 \right] -\frac{1}{2 \sigma_w^2} \sum_{i=1}^n w_i \left( \phi^{\lambda, x_0, s}(x_i) - \mu_w \right)^2 \\
& - \left( \sum_{i=1}^n w_i \right) \log s + (\lambda - 1) \sum_{i=1}^n w_i \sgn(x_i - x_0) \log \left[1 + \frac{|x_i - x_0|}{s} \right]
\end{split}
\end{equation}
```
where $\mu_w$ and $\sigma^2_w$ are the weighted mean and weighted variance of the Yeo-Johnson transformed feature $\phi_{\text{YJ}}^{\lambda, x_0, s} (\mathbf{X})$:

# Appendix B: Optimisation of transformation parameters

@@

Let $\mathbf{X}$ be a feature with ordered feature values, and $\mathbf{Y}^\lambda =\psi^{\lambda} \left(\mathbf{X} \right)$ and $\mathbf{Y}^{\lambda, x_0} =\psi^{\lambda, x_0} \left(\mathbf{X} \right)$ its transformed values using conventional and shift and scale invariant power transformations, respectively. Since power transformations are monotonic, $\mathbf{Y}$ will likewise be ordered.

Below we will focus on criteria based on the empirical density function and those based on skewness and kurtosis of the transformed featured. Other potential criteria, such as the Shapiro-Wilk test statistic [@Shapiro1965-zd] are not investigated here. In the case of the Shapiro-Wilk test statistic this is because of lack of scalability to features with many ($> 5000$) instances, and because adapting the test statistic to include weights is not straightforward.
